Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-06 paper/paper.tex
Proofreading

\Citet{van2012systematic,van2010abstracting} proposed the Abstracting Abstract Machines (AAM)
methodology which provides a recipe for constructing sound abstract interpreters
for higher-order functional languages from concrete abstract machines.
%~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine that models a store 
(e.g. the CESK machine, Krivine's machine, etc.), if we allocate continuations 
in the store and draw all allocated addresses (both for regular values and 
for continuations) from a finite set, then
we obtain an abstract interpreter with a finite state space, which can be used for 
performing sound static analysis and is guaranteed to terminate.
By using different address allocators \cite{Gilray:2016:ACP:2951913.2951936},
one can further instantiate different polyvariant control-flow analyses.

@@

The CESK machine is an abstract machine for describing the semantics of and evaluating
$\lambda$-calculus \cite{felleisen1987calculus}. The CESK machine models program execution
as state transitions in a small-step fashion. As its name suggests, a machine state has
four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the addresses of all variables in the lexical scope.
3) \textit{Store} models the heap of a program as a map from addresses to values.
  The address space consists of numbers (0-indexed).
  In our toy language, the only category of value is a closure, i.e., a function paired with
  an environment.
4) \textit{Continuation} represents the program's execution context. In this paper, we instantiate the 
execution context as a call stack consisting of a list of frames.

@@

\subsubsection{Single-Step Transition}
Before describing how the machine evaluates expressions, we must first define several helper
functions. As mentioned in Section~\ref{anfsyntax}, atomic expressions are either a variable
or a literal @lambda@ term. As such, the atomic expression evaluator @atomicEval@ handles
these two cases and evaluates atomic expressions to closures in a straightforward way.
The @alloc@ function generates a fresh address, and always allocates an unique integer
in the domain of @store@.
The @isAtomic@ function is used as a predicate to determine if the expression is atomic.

@@

We can now faithfully describe the state transition function @step@,
which when given a machine state, determines its successor state.
The function @step@ is a partial function that only handles non-final states,
which must have a successor; the final case of a state is handled by function
@drive@, which will be mentioned in Section~\ref{sec:drive}.

@@

As shown and previously discussed, we examine the only two non-final cases which the state
may be.

@@

In this section, we derive the abstract abstract machine from the concrete
CESK machine, and also show how to instantiate useful $k$-call-sensitive 
control-flow analysis.

@@

Similar to the CESK machine, the machine state of AAM has a control, an environment, a store,
and continuation, as well as a timestamp. However, there are several notable differences
between AAM's store and CESK machine's store. In AAM, the store maps addresses to sets of
values; it stores all possible values for a particular address. As such, dereferencing
addresses becomes nondeterministic. Also, the store performs \emph{joining}, rather than
overwriting, when updating elements. Furthermore, the continuations are likewise allocated
on the store instead of formed into a linked list, and the continuation component becomes
an address that maps to a set of continuations in the store instead of directly embedded
in the state.

@@

Similar to the CESK machine, to run (analyze) a program we first use the @inject@ function
to construct the initial state given to the program. Note that the initial continuation store
has a built-in mapping that maps the continuation address for @Halt@ to an empty set of 
continuations. We also provide an empty program contour as our initial time.

@@

However, in contrast to the concrete CESK machine, the @drive@ function performs 
collecting semantics instead of the valuation semantics. That is, for the purpose of
analyzing programs, the function @drive@ collects all the intermediate machine states 
as the program is abstractly executing. The following code shows a variant of the 
work-list algorithm to find the fixed-point of the set of states.
Function @drive@ always applies function @step@ to the head element @hd@ of
the work-list @todo@ if @hd@ is unseen. It then inserts the result of @step@ to
the rest of work-list, and in the meantime adds @hd@ to the explored states set.
If the work-list is empty, @drive@ simply returns the set of reachable states
up to the current execution point.

@@

In this section, we describe a variant of AAM that allows the stack to be unbounded 
which uses a precise model of the call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call structure as the same as at run-time
(so-called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

The pushdown A\textsuperscript{2}M with an unbounded stack is also described by \citeauthor{van2012systematic} 
\cite{van2012systematic}. 
The reason we show it here is to keep the consistency of abstract semantics during 
transformations -- since the eventual abstract definitional interpreters simply inherit
a precise control-flow from its defining language, we also would like to start from a
pushdown A\textsuperscript{2}M that is also able to precisely match calls and returns.
For readers who are not familiar with pushdown analysis, we have a detailed discussion 
in Section~\ref{pdcfarevisit}.

@@

The state transition function @step@ shown below is still nondeterministic, but the only 
nondeterminism happening is when dereferencing the callee @f@ from the function application 
@App(f, ae)@.

@@

\textit{Note on Termination.}
Unfortunately, even though other components in the states are finite, this A\textsuperscript{2}M with an unbounded
stack still may diverge when analyzing some programs.
This is because the unbounded stack can grow to an arbitrary depth which implies that the state
space is possibly infinite; the analysis may therefore not terminate for all programs if we
simply enumerate reachable states. To see this, consider a program that has two mutually
recursive functions:

@@

Function @f1@ and @f2@ mutually invoke each other, so the stack will alternately pushing
frames @f1@ and @f2@ onto the top of the current stack. However, no two existing stack
components are identical in the state space. One can decide the reachability of a state
through Dyck state graph \cite{earl2010pushdown, earl2012introspective}, but simply
caching the explored states (i.e. @seen@) would not always make the analyzer terminate.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match. Now we begin describing the
transformations step by step. Our base machine is the A\textsuperscript{2}M with an unbounded
stack, as we will later transform the stack structure to higher-order functions
representing continuations, it does not matter what kind of A\textsuperscript{2}M we start from.
This is because it would not be possible to construct a new frame on the stack, nor
to allocate continuations in the store after Refunctionalization.
Thus, our choice to start from an unbounded-stack A\textsuperscript{2}M is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

In his paper \textit{Defunctionalized Interpreters for Programming Languages},
Danvy observes that for deterministic languages, a reduction semantics can be seen as 
a structural operational semantics in continuation-passing style, where the reduction context is
a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}. Therefore
refunctionalizing such a reduction semantics for a deterministic language yields 
an interpreter in continuation-passing style. 
Nevertheless, doing so for nondeterministic languages yields an interpreter
with multiple layers of continuation \cite{Danvy:2006:RW:2171265.2171268, DANVY2009534},
where the additional layer of continuation represents the nondeterministic choices.

We apply this idea to AAM and our first step of transformations is to linearize all
nondeterministic choices. This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of control, and define a case class @NDCont@
that saves the information at a fork point when we have nondeterministic target closures.
We also add a new field @ndk@ of type @NDCont@ to the definition of state, which we now 
call @NDState@. For clarity of presentation in differentiating between the two continuations, 
we elect to call the first a \emph{continuation}, and the second a \emph{meta-continuation}
\footnote{These are also called \emph{success continuation} and \emph{failure
continuation} \cite{Danvy:1990:AC:91556.91622}.}.

@@

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the target function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added to convert itself to type @State@.  
Function @step@ now becomes of type @NDState => Option[NDState]@. Returning a @None@ value means
reaching the halt.

@@

If the continuation @konts@ is empty, then the machine has reached an end of the computation;
otherwise, we should return the values of @ae@ to its caller which is contained in the top
frame of the stack. However, since a meta-continuation @ndk@ is introduced, we need to dispatch
more cases when @konts@ is empty:

\begin{itemize}
  \item If the meta-continuation @ndk@ is also empty, then the machine has reached the end of \textit{all} 
    computations. Besides, the control component @ae@ now is an atomic value, therefore @None@ is returned.
  \item If the target closures set of the front-most @NDCont@ is empty (@Nil@ in the code),
    then we have computed all the closures at this fork point and should move to the next fork 
    point. Since the meta-continuation @nkd@ is represented by a list, and we always append new 
    elements to its front, we resume to the most recent fork point by popping up the front-most 
    element and using the tail of that list.
  \item Otherwise, we pop a closure from the set, and build a new environment and store for 
    evaluating the body expression of that closure. We use the same frame list and time which 
    are copied from the fork point rather than current one.
    The meta-continuation @new_ndk@ is updated by the removal of that popped closure, which makes
    sure the closure set will eventually shrink to empty.
\end{itemize}

@@

The @drive@ function is also changed: there is no work-list anymore, as all the potentially 
unexplored states are embedded into the continuations.
The termination of the analysis occurs when the value @None@ is returned from calling @step@.
Otherwise, we extract the next state @nnds@ and recur on function @drive@.
The set of explored states is preserved as before.

At this point, we have obtained a linearized abstract abstract machine. 
Intuitively, if we imagine that the classical AAM explores a graph of reachable states, 
then the linearized A\textsuperscript{2}M flattens the graph in depth-first order to a linear sequence of states.
It has been observed that ``\textit{failure can be replaced by a list of successes}''
\cite{10.1007/3-540-15975-4_33} and that ``\textit{one programmer’s lazy-list 
constructor is another programmer’s success and failure continuations}'' \cite{danvy2006a}; 
the @step@ function in AAM is the list constructor used in a lazy way, and here we turn it 
into an explicit data types representing continuation.
The inverse transformation of this step is to apply \citeauthor{10.1007/3-540-15975-4_33}'s 
technique that converts the data types representing meta-continuations (failure continuations) 
back to lazy list representation, where the elements in the list are plain machine states.

@@

Then we apply a lightweight fusion transformation that combines the @step@ and @drive@
functions into a single function.
The fused function @drive_step@ takes an @NDState@ and a set of explored states @seen@ 
as arguments and returns a set of reachable states once it terminates.
It is essentially formed by merging the functionality of @step@ into the @drive@ function.
Since each case in function @step@ is statically known, we can inline the pattern match on the
@Option@ type and replace it with returning @new_seen@ when @None@ is matched, or calling 
@drive_step@ recursively when @Some@ is matched.

@@

Function @continue@ is identical to what we have seen for the second case in @drive_step@,
except that the dispatching logic for empty @konts@ is lifted to a separate function 
@ndcontinue@.
It matches on the list representation of context: An empty list @Nil@ represents the
halt of a computation path, whereas a cons @::@ means the top frame of the list
holds the innermost context.

@@

Function @ndcontinue@ is identical to the dispatching logic in function @continue@ when @konts@ is empty.
When @ndk@ is empty, the current explored states @seen@ is returned; the analysis then terminates.
Otherwise, a new @NDState@ is constructed depending on the status of the closures set.
% first call to drive_step can be replaced with continue. ae is atomic, konts is empty.

The purpose of disentanglement is to reveal the defunctionalized form of AAM.
Functions @continue@ and @ndcontinue@ that dispatches different data types representing 
continuations are disentangled from @drive_step@. 
The inverse transformation of this step would be simply inlining @ndcontinue@ into @continue@,
and then inlining @continue@ into @drive_step@.

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain an abstract interpreter written in extended
continuation-passing style\cite{Danvy:1990:AC:91556.91622}.
This transformation can be also regarded as making the control flows of the defined 
language be part of the control flows of the meta-language.
In this section, we first present a direct refunctionalization of the continuations.
Next, we adopt a caching algorithm to obtain a computable analysis\cite{darais2017abstracting},
and further prepare the interpreter for transforming to direct-style later.

@@

By looking at the definition of the first-order function @continue@ from the last section,
we may find it calls the meta-continuation @ndcontinue@ inside of its body.
This observation leads us to define the types of higher-order functions representing 
continuation and meta-continuation:

@@

The continuation @Cont@ as a higher-order function takes a configuration, a set of explored
configurations, as well as a meta-continuation as its arguments. While the meta-continuation
@MCont@ only takes a configuration and a set of explored configurations as arguments.
Now, the skeleton of function @aeval@ looks as follows:

@@

The continuation and meta-continuation are added as arguments to @aeval@.
Before jumping into the details of refunctionalizing the stack structure and constructing
@new_cont@ and @new_mcont@, we have two important observations in the disentangled A\textsuperscript{2}M
where the continuations were represented by data types:
1) When constructing the @new_ndk@, we observe that it explicitly uses @new_frames@ as one
argument of its top @NDCont@ object. The higher-order version of these continuations follows
the same pattern.
2) In the meta-continuation dispatching function @ndcontinue@, the size of the closures set is
decreasing every time while the other parts of meta-continuation are kept. This means in 
the higher-order version, we also need an ``ability'' to update the top frame of meta-continuation
and keep the rest as the same.

Let us first look into @new_cont@. The body of @new_cont@ is similar to the first-order dispatching
function @continue@. The invocation of @aeval@ uses the latest meta-continuation @m@ instead of the outer
one, @mcontinue@. Note that the new environment @new_env@ is updated based on the outer environment;
while the atomic evaluation uses the latest environment @env_@.

@@

Then to address the circularity of the meta-continuations, we define an auxiliary function
@makeMCont@ to construct @new_mcont@, which takes a list of closures as the argument and returns 
the meta-continuation. 
Function @makeMCont@ is a recursive function where the recursive invocation happens when 
constructing the meta-continuation based on the rest of closures set. 
Now we can dynamically generate the meta-continuation inside of a meta-continuation.
Meanwhile, the @new_cont@ we defined above is used here, simply because of different closures 
actually share the same continuation.
The case where the empty closures set occurs is also interesting. At this case,
the outer meta-continuation @mcontinue@ is returned.
It would be clear if we inductively think the base case where the closures set contains only one 
closure, then @makeMCont@ is invoked with an empty set, therefore should be just the 
current meta-continuation @mcontinue@.

@@

One last very important thing is about the initial continuation and meta-continuation.
The initial continuation will be invoked when one computation path reaches its end,
so we invoke the meta-continuation @m@ to continue exploring other paths.
The initial meta-continuation will be invoked when all computation paths reach their ends,
so we terminate with the explored configurations @seen@. They correspond to the 
cases where @konts@ is empty and @konts@ and @ndk@ are both empty, respectively.

@@

Now we have completely refunctionalized an abstract abstract machine; the two layers
of continuations are represented by higher-order functions. The inverse transformation 
of this step is known as \textit{de}functionalization, that converts the higher-order
representing continuations back to first-order data types representation 
\cite{Danvy:2001:DW:773184.773202, Danvy:2008:DIP:1411204.1411206}.

@@

\begin{itemize}
  \item The components of the configuration are lifted as arguments of the evaluation
    function.
  \item The refunctionalized A\textsuperscript{2}M collects all possible configurations, but our final target is an 
    abstract definitional interpreter that cares about final evaluation results.
    So starting from this section our abstract semantic artifact returns a set of final 
    values instead of collected configurations. 
  \item The configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible values and reachable paths.
  \item The function @aeval@ is simplified by introducing and using a nondeterministic 
    operator @nd@ which makes @aeval@ only uses one layer of continuation.
\end{itemize}

We will begin by discussing these changes separately, and finally see how they work together.

@@

\subsubsection{Cache-Passing Style}
% why cache? problem of set of states.
Recall that the latent assumption of the work-list algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.
The case class @Cache@ and its operations are defined as follow:

@@

As previously mentioned, the components of configurations are lifted as arguments to
@aeval@. Given the cache and the simplified one-layer continuation (as we will see later), 
the shape of function @aeval@ now looks like:

@@

%TODO: co-inductive
Upon entering @aeval@, we first determine whether the @out@ cache contains some values 
for the current configuration and use them immediately (if present) by calling @continue@.
This corresponds to small-step AAM with a work-list: when we have seen a state,
we can simply discard the state and continue working through the rest of the work-list.

@@

We observe that in the refunctionalized A\textsuperscript{2}M, @makeMCont@ is invoked with the tail part of
the closure set, and it actually repeats the code that creates the new environment and 
store for the body expression of a closure.
Based on observation, we present a nondeterministic operator that abstracts this pattern:

@@

Given a collection of elements of type @T@, an initial accumulator of type @Ans@,
a function @k@ that works on elements of type @T@, @nd@ chooses an element in
the collection and applies @k@ to it iteratively, and accumulates the values that
returned from @k@ when recurring on the rest of the collection. Eventually, @nd@
returns the accumulated value @acc@ when there is no element in the collection.
The latest accumulated cache @acc.cache@ is exposed to @k@ when applied.

\subsubsection{Put Them Together}

With the skeleton of cache-and-continuation-passing style @aeval@ in mind, we may now
use the @nd@ operator to fill in the details. The first case of match is that @e@
is a @let@ expression:

>>>>

\Citet{van2012systematic,van2010abstracting} proposed the Abstracting abstract machines (AAM)
methodology which provides a recipe for constructing sound abstract interpreters
for higher-order functional languages from concrete abstract machines.
%~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine that models a store 
(e.g. the CESK machine, Krivine's machine, etc.), if we allocate continuations 
in the store and draw all allocated addresses (both for regular values and 
for continuations) from a finite set, then
we obtain an abstract interpreter with a finite state space, which can be used for 
performing sound static analysis and is guaranteed to terminate.
By using different address allocators \cite{Gilray:2016:ACP:2951913.2951936},
one can further instantiate different polyvariant control-flow analyses.

@@

The CESK machine is an abstract machine for describing the semantics of and evaluating
a $\lambda$-calculus \cite{felleisen1987calculus}. The CESK machine models program execution
as state transitions in a small-step fashion. As its name suggests, a machine state has
four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the addresses of all variables in the lexical scope.
3) \textit{Store} models the heap of a program as a map from addresses to values.
  The address space consists of numbers (0-indexed).
  In our toy language, the only category of value is a closure, i.e., a function paired with
  an environment.
4) \textit{Continuation} represents the program's execution context. In this paper, we instantiate the 
execution context as a call stack consisting of a list of frames.

@@

\subsubsection{Single-Step Transition}
Before describing how the machine evaluates expressions, we must first define several helper
functions. As mentioned in Section~\ref{anfsyntax}, atomic expressions are either a variable
or a literal @lambda@ term. As such, the atomic expression evaluator @atomicEval@ handles
these two cases and evaluates atomic expressions to closures in a straightforward way.
The @alloc@ function generates a fresh address, and always allocates a unique integer
in the domain of @store@.
The @isAtomic@ function is used as a predicate to determine if the expression is atomic.

@@

We can now faithfully describe the state transition function @step@,
which when given a machine state, determines its successor state.
The function @step@ is a partial function that only handles non-final states,
which must have a successor; the final case of a state is handled by function
@drive@, which will be explained in Section~\ref{sec:drive}.

@@

As shown and previously discussed, we examine the only two non-final cases which the state
may be:

@@

In this section, we derive the respective abstract abstract machine from the concrete
CESK machine, and also show how to instantiate useful $k$-call-sensitive 
control-flow analysis.

@@

Similar to the CESK machine, the machine state of an A\textsuperscript{2}M has a control, an environment, a store,
and continuation, as well as a timestamp. However, there are several notable differences
between an A\textsuperscript{2}M's store and the CESK machine's store. In AAM, the store maps addresses to sets of
values; it stores all possible values for a particular address. As such, dereferencing
addresses becomes nondeterministic. Also, the store performs \emph{joining}, rather than
overwriting, when updating elements. Furthermore, the continuations are likewise allocated
on the store instead of formed into a linked list, and the continuation component becomes
an address that maps to a set of continuations in the store instead of directly embedded
in the state.

@@

Similar to the CESK machine, to run (analyze) a program, we first use the @inject@ function
to construct the initial state given to the program. Note that the initial continuation store
has a built-in mapping that maps the continuation address for @Halt@ to an empty set of 
continuations. We also provide an empty program contour as our initial time.

@@

However, in contrast to the concrete CESK machine, the @drive@ function performs 
collecting semantics instead of valuation semantics. That is, for the purpose of
analyzing programs, the function @drive@ collects all the intermediate machine states 
as the program is abstractly executing. The following code shows a variant of the 
work-list algorithm to find the fixed-point of the set of states.
Function @drive@ always applies function @step@ to the head element @hd@ of
the work-list @todo@ if @hd@ is unseen. It then inserts the result of @step@ to
the rest of work-list, and in the meantime adds @hd@ to the explored states set.
If the work-list is empty, @drive@ simply returns the set of reachable states
up to the current execution point.

@@

In this section, we describe a variant of AAM that allows the stack to be unbounded,
which uses a precise model of the call stack as in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call structure as the same as at run-time
(so-called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

The pushdown AAM with an unbounded stack is also described by \citeauthor{van2012systematic} 
\cite{van2012systematic}. 
We show it here is to maintain consistency of abstract semantics during
transformations -- since the eventual abstract definitional interpreters simply inherit
a precise control-flow from their defining language, we also would like to start from a
pushdown A\textsuperscript{2}M that is also able to precisely match calls and returns.
For readers who are not familiar with pushdown analysis, we provide a detailed discussion
in Section~\ref{pdcfarevisit}.

@@

The state transition function @step@ shown below is still
nondeterministic, but only when dereferencing the callee @f@ from the
function application @App(f, ae)@.

@@

\textit{Note on Termination.}
Unfortunately, even though other components in the states are finite, this version of AAM with an unbounded
stack may still diverge when analyzing some programs.
This is because the unbounded stack can grow to an arbitrary depth which implies that the state
space is possibly infinite; the analysis may therefore not terminate for all programs if we
simply enumerate reachable states. To see this, consider a program that has two mutually
recursive functions:

@@

Function @f1@ and @f2@ mutually invoke each other, so the stack will alternately push
frames @f1@ and @f2@ onto the top of the current stack. However, no two existing stack
components are identical in the state space. One can decide the reachability of a state
through Dyck state graph \cite{earl2010pushdown, earl2012introspective}, but simply
caching the explored states (i.e. @seen@) would not always make the analyzer terminate.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match. Now we begin describing the
transformations step by step. Our base machine is the A\textsuperscript{2}M with an unbounded
stack; as we will later transform the stack structure to higher-order functions
representing continuations, it does not matter what kind of A\textsuperscript{2}M we start from.
This is because it would not be possible to construct a new frame on the stack, nor
to allocate continuations in the store after refunctionalization.
Thus, our choice to start from an unbounded-stack A\textsuperscript{2}M is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

In his paper \textit{Defunctionalized Interpreters for Programming Languages},
Danvy observes that for deterministic languages, a reduction semantics can be seen as 
a structural operational semantics in continuation-passing style, where the reduction context is
a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}. Therefore,
refunctionalizing such a reduction semantics for a deterministic language yields 
an interpreter in continuation-passing style. 
Nevertheless, doing so for nondeterministic languages yields an interpreter
with multiple layers of continuation \cite{Danvy:2006:RW:2171265.2171268, DANVY2009534},
where the additional layer of continuation represents the nondeterministic choices.

We apply this idea to AAM, and thus our first transformation step is to linearize all
nondeterministic choices. This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of control, and define a case class @NDCont@
that saves the information at a fork point when we have nondeterministic target closures.
We also add a new field @ndk@ of type @NDCont@ to the definition of state, which we now 
call @NDState@. For clarity of presentation in differentiating between the two continuations, 
we elect to call the first a \emph{continuation}, and the second a \emph{meta-continuation}
\footnote{These are also called \emph{success continuation} and \emph{failure
continuation} \cite{Danvy:1990:AC:91556.91622}.}.

@@

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the target function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added to convert itself to type @State@.  
@step@ now becomes of type @NDState => Option[NDState]@. Returning a @None@ value means
reaching the halt.

@@

If the continuation @konts@ is empty, then the machine has reached an end of the computation.
Otherwise, we should return the values of @ae@ to its caller which is contained in the top
frame of the stack. However, since a meta-continuation @ndk@ is introduced, we need to dispatch
more cases when @konts@ is empty:

\begin{itemize}
  \item If the meta-continuation @ndk@ is also empty, then the machine has reached the end of \textit{all} 
    computations. Besides, the control component @ae@ now is an atomic value: @None@ is returned.
  \item If the target closures set of the front-most @NDCont@ is empty (@Nil@ in the code),
    then we have computed all the closures at this fork point and should move to the next fork 
    point. Since the meta-continuation @nkd@ is represented by a list, and we always append new 
    elements to its front, we resume to the most recent fork point by popping the front-most 
    element and using the tail of that list.
  \item Otherwise, we pop a closure from the set, and build a new environment and store for 
    evaluating the body expression of that closure. We use the same frame list and time which 
    are copied from the fork point rather than current one.
    The meta-continuation @new_ndk@ is updated by the removal of that popped closure, which guarantees
    the closure set will eventually shrink to empty.
\end{itemize}

@@

The @drive@ function is also changed: there is no work-list anymore, as all the potentially 
unexplored states are embedded into the continuations.
The termination of the analysis occurs when the value @None@ is returned from calling @step@.
Otherwise, we extract the next state @nnds@ and recurse on function @drive@.
The set of explored states is preserved as before.

At this point, we have obtained a linearized abstract abstract machine. 
Intuitively, if we imagine that the classical AAM explores a graph of reachable states, 
then the linearized version of AAM flattens the graph in depth-first order to a linear sequence of states.
It has been observed that ``\textit{failure can be replaced by a list of successes}''
\cite{10.1007/3-540-15975-4_33} and that ``\textit{one programmer’s lazy-list 
constructor is another programmer’s success and failure continuations}'' \cite{danvy2006a}; 
the @step@ function in AAM is the list constructor used in a lazy way, and here we turn it 
into an explicit data types representing continuation.
The inverse transformation of this step is to apply \citeauthor{10.1007/3-540-15975-4_33}'s 
technique that converts the data types representing meta-continuations (failure continuations) 
back to lazy list representation, where the elements in the list are plain machine states.

@@

We next apply a lightweight fusion transformation that combines the @step@ and @drive@
functions into a single function.
The fused function @drive_step@ takes an @NDState@ and a set of explored states @seen@ 
as arguments and returns a set of reachable states once it terminates.
It is essentially formed by merging the functionality of @step@ into the @drive@ function.
Since each case in function @step@ is statically known, we can inline the pattern matching on the
@Option@ type and replace it with returning @new_seen@ when @None@ is matched, or by calling 
@drive_step@ recursively when @Some@ is matched.

@@

Function @continue@ is identical to what we have seen for the second case in @drive_step@,
except that the dispatching logic for empty @konts@ is lifted to a separate function 
@ndcontinue@.
It matches on the list representation of context: an empty list @Nil@ represents the
halt of a computation path, whereas a cons @::@ means the top frame of the list
holds the innermost context.

@@

Function @ndcontinue@ is identical to the dispatching logic in @continue@ when @konts@ is empty.
When @ndk@ is empty, the current set of explored states @seen@ is returned, after which the analysis terminates.
Otherwise, a new @NDState@ is constructed depending on the status of the closures set.
% first call to drive_step can be replaced with continue. ae is atomic, konts is empty.

The purpose of disentanglement is to reveal the defunctionalized form of our A\textsuperscript{2}M.
Functions @continue@ and @ndcontinue@ that dispatch different data types representing 
continuations are disentangled from @drive_step@. 
The inverse transformation of this step would be simply inlining @ndcontinue@ into @continue@,
and then inlining @continue@ into @drive_step@.

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain an abstract interpreter written in extended
continuation-passing style~\cite{Danvy:1990:AC:91556.91622}.
This transformation can be also regarded as making the control flows of the defined 
language be part of the control flows of the meta-language.
In this section, we first present a direct refunctionalization of the continuations.
Next, we adopt a caching algorithm to obtain a computable analysis~\cite{darais2017abstracting},
and further prepare the interpreter for transformation to direct-style at a later point.

@@

By looking at the definition of the first-order function @continue@ from the last section,
we may find it calls the meta-continuation @ndcontinue@ inside of its body.
This observation leads us to define the types of higher-order functions representing
continuations and meta-continuations:

@@

The continuation @Cont@ is a higher-order function which takes a configuration, a set of explored
configurations, as well as a meta-continuation as its arguments, while the meta-continuation
@MCont@ only takes a configuration and a set of explored configurations as arguments.
Now, the skeleton of function @aeval@ looks as follows:

@@

The continuation and meta-continuation are added as arguments to @aeval@.
Before jumping into the details of refunctionalizing the stack structure and constructing
@new_cont@ and @new_mcont@, we have two important observations in the disentangled A\textsuperscript{2}M
where the continuations were represented by data types:
1) When constructing the @new_ndk@, we observe that it explicitly uses @new_frames@ as one
argument of its top @NDCont@ object. The higher-order version of these continuations follows
the same pattern.
2) In the meta-continuation dispatching function @ndcontinue@, the size of the closures set is
decreasing every time while the other parts of meta-continuation are kept. This means in 
the higher-order version, we also need an ability to update the top frame of meta-continuation
and keep the rest as the same.

Let us first look into @new_cont@. The body of @new_cont@ is similar to the first-order dispatching
function @continue@. The invocation of @aeval@ uses the latest
meta-continuation @m@ instead of the outer @mcontinue@. Note that the
new environment @new_env@ is updated based on the outer environment;
the atomic evaluation uses the latest environment @env_@.

@@

To address the circularity of the meta-continuations, we define an auxiliary function
@makeMCont@ to construct @new_mcont@, which takes a list of closures as the argument and returns the meta-continuation.
Function @makeMCont@ is a recursive function where the recursive invocation happens when 
constructing the meta-continuation based on the rest of closures set. 
Now we can dynamically generate the meta-continuation inside of a meta-continuation.
Meanwhile, the @new_cont@ we defined above is used here, simply because of different closures 
actually share the same continuation.
The case where the empty closures set occurs is also interesting. At this case,
the outer meta-continuation @mcontinue@ is returned.
This becomes clear thinking inductively; our base case is such that the closures set contains only one 
closure. @makeMCont@ is thus invoked with an empty set, and should therefore be just the
current meta-continuation @mcontinue@.

@@

We present one final, though very important, note regarding the initial continuation and meta-continuation.
The initial continuation will be invoked when one computation path reaches its end,
so we invoke the meta-continuation @m@ to continue exploring other paths.
The initial meta-continuation will be invoked when all computation paths reach their ends,
so we terminate with the explored configurations @seen@. They correspond to the 
cases where @konts@ is empty and @konts@ and @ndk@ are both empty, respectively.

@@

We have now completely refunctionalized an abstract abstract machine; the two layers
of continuations are represented by higher-order functions. The inverse transformation
of this step is known as \textit{de}functionalization, which converts the higher-order functions
representing continuations back to first-order data
types~\cite{Danvy:2001:DW:773184.773202, Danvy:2008:DIP:1411204.1411206}.

@@

\begin{itemize}
  \item The components of the configuration are lifted as arguments of the evaluation
    function.
  \item The refunctionalized AAM collects all possible configurations, but our final target is an 
    abstract definitional interpreter that cares about final evaluation results.
    As such, starting from this section, our abstract semantic artifact returns a set of final 
    values instead of collected configurations. 
  \item The configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible values and reachable paths.
  \item The function @aeval@ is simplified by introducing and using a nondeterministic 
    operator @nd@ which makes @aeval@ only uses one layer of continuation.
\end{itemize}

We begin discussing these changes separately, and finally see how they work together.

@@

\subsubsection{Cache-Passing Style}
% why cache? problem of set of states.
Recall that the latent assumption of the work-list algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption, but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.
The case class @Cache@ and its operations are defined as follow:

@@

As previously mentioned, the components of configurations are lifted as arguments to
@aeval@. Given the cache and the simplified one-layer continuation (as we will see later),
the shape of function @aeval@ now looks as follows:

@@

%TODO: co-inductive
Upon entering @aeval@, we first determine whether the @out@ cache contains some values 
for the current configuration, and (if present) use them immediately by
calling @continue@.
This corresponds to small-step AAM with a work-list: when we have seen a state,
we can simply discard the state and continue working through the rest of the work-list.

@@

We observe that in the refunctionalized AAM, @makeMCont@ is invoked with the tail part of
the closure set, and actually repeats the code that creates the new environment and
store for the body expression of a closure.
Based on this observation, we present a nondeterministic operator that abstracts this pattern:

@@

Given a collection of elements of type @T@, an initial accumulator of
type @Ans@, and a function @k@ that works on elements of type @T@, @nd@
chooses an element in
the collection and applies @k@ to it iteratively, and accumulates the values that
returned from @k@ when recurring on the rest of the collection. Eventually, @nd@
returns the accumulated value @acc@ when there is no element in the collection.
The latest accumulated cache @acc.cache@ is exposed to @k@ when applied.

\subsubsection{Putting Them Together}

With the skeleton of cache-and-continuation-passing style @aeval@ in mind, we may now
use the @nd@ operator to fill in the details. The first case of our
match is that @e@ is a @let@ expression:
